% !TeX root = ../main.tex

\section{Introduction}
\label{sec:introduction}

Large language models have become the default substrate for a widening range of
cognitive tasks: planning, tool use, causal reasoning, memory management, and
uncertainty calibration.  Yet the transformer architecture was designed for
sequence-to-sequence transduction, not for any of these tasks.  The prevailing
approach---encoding all context, instructions, and state into a text
prompt---conflates two fundamentally different concerns: \emph{what to say}
(surface-form generation) and \emph{what to think} (world modelling,
inference, control).

This paper proposes an alternative division of labour.  We treat the LLM as a
\textbf{language organ}: a frozen decoder whose only responsibility is to
produce fluent, well-formed language.  All cognitive operations---persistent
memory, causal reasoning, decision-making under uncertainty, belief revision,
and top-down control---are handled by a separate \textbf{cognitive substrate}
that communicates with the LLM exclusively through \emph{grafts}: small modules
that modify the residual stream, lexical plan, and logit distribution at every
decoding step, without touching the frozen model's weights and without
consuming prompt tokens.

The substrate is assembled from components with published mathematical
guarantees rather than from learned or heuristic modules:

\begin{itemize}[nosep]
  \item \textbf{Active inference} (Friston): POMDPs that minimise Expected Free
    Energy, coupling epistemic and pragmatic value in a single decision
    criterion.
  \item \textbf{Finite SCM with do-calculus} (Pearl): exact interventional and
    counterfactual queries, with automatic backdoor and frontdoor
    identification.
  \item \textbf{Holographic Reduced Representations} (Plate): compositional
    bind/unbind/bundle algebra in $O(d \log d)$ via FFT, with capacity
    $\sim 0.5\,d / \log d$.
  \item \textbf{Modern Continuous Hopfield networks} (Ramsauer et al., 2020):
    one-step retrieval with exponential storage capacity in $d$.
  \item \textbf{Split-conformal prediction} (Vovk; Romano et al., 2020):
    distribution-free coverage guarantee $P[y \in C(x)] \geq 1 - \alpha$.
  \item \textbf{Multivariate Hawkes processes}: self- and mutually-exciting
    event streams for working-memory ``heat.''
  \item \textbf{Predictive coding}: cross-entropy gap between graft-on and
    graft-off forwards, yielding the substrate's surprise signal.
\end{itemize}

The LLM's weights are \emph{never updated}.  Behavioural changes arise entirely
from the substrate's reads and writes through grafts.  This design has three
practical advantages: (1)~the language organ can be swapped (e.g.\ Llama~3.2
$\to$ a future checkpoint) without retraining any cognitive component;
(2)~the substrate's state is fully persistent (SQLite WAL), surviving restarts
and enabling long-running agents; and (3)~every cognitive claim---retrieval
correctness, coverage, capacity, convergence---has a paper trail rather than a
vibe.

\paragraph{Contributions.}
\begin{enumerate}[nosep]
  \item The \textsc{Mosaic} architecture and its graft-based communication
    protocol between a frozen LLM and a persistent cognitive substrate
    (\S\ref{sec:methods}).
  \item A three-tier evaluation framework: standard NLP benchmarks for
    regression testing, architecture probes for graft effectiveness, and
    substrate-specific benchmarks for algebraic guarantees
    (\S\ref{sec:experiments}).
  \item A fully automated benchmark-to-paper pipeline that regenerates all
    experiment tables, figures, and prose from fresh benchmark runs, closing
    the loop from code change to typeset report.
\end{enumerate}
