% !TeX root = ../main.tex

\section{Methods}
\label{sec:methods}

\subsection{Architecture Overview}

The \textsc{Mosaic} system consists of two interacting systems, inspired by dual-process
theories of cognition \cite{kahneman2011thinking}:

\begin{itemize}[nosep]
  \item \textbf{System~1 (Language Organ):} A frozen causal language model
    (Llama~3.2 1B Instruct checkpoint) that produces fluent surface forms.  Its
    weights are never modified.
  \item \textbf{System~2 (Cognitive Substrate):} A persistent collection of
    algebraic and probabilistic modules---memory, world model, agents, algebra,
    and control---that runs inference, maintains state, and biases the language
    organ through grafts.
\end{itemize}

Communication flows exclusively through three \emph{graft slots} in the frozen
model's forward pass.  Throughout we refer to the graft-augmented frozen LM arm as the
\emph{Broca shell} (sometimes ``Broca stack''): it is the surface decoder paired with this substrate inside \textsc{Mosaic}, not a separate system.

\begin{enumerate}[nosep]
  \item \textbf{Residual-stream graft} (\texttt{final\_hidden}): Projects a
    continuous cognitive-frame vector into the host's residual stream at a
    target signal-to-noise ratio calibrated to win over an arbitrarily deep
    autoregressive prefix.
  \item \textbf{Lexical-plan graft} (\texttt{final\_hidden}): Biases the
    next-token distribution toward a small ``speech plan'' of tokens the
    substrate would like the host to emit, with confidence-modulated decay.
  \item \textbf{Logit-bias graft} (\texttt{logits}): Content-aware token bias
    derived from the cognitive frame's subject, predicate, and answer, encoded
    into subwords.
\end{enumerate}

\subsection{Cognitive Substrate Components}

\paragraph{Persistent semantic memory.}
All symbolic knowledge is stored in SQLite with WAL journaling.  The memory
supports subject--predicate--object triples with confidence scores, an
append-only claims ledger, and a reflections table that records every belief
revision decision with full provenance.

\paragraph{Active inference.}
Two categorical POMDPs run in parallel: a spatial/action POMDP (Tiger-door
problem) and a causal epistemic POMDP derived from the structural causal model.
A \texttt{CoupledEFEAgent} arbitrates by comparing the minimal one-step Expected
Free Energy of each faculty.  The winning posterior entropy modulates the LLM's
sampling temperature: a confused substrate produces exploratory decoding; a
decisive substrate produces near-greedy decoding.

\paragraph{Structural causal model.}
A finite SCM (\texttt{FiniteSCM}) supports exact $\doop(\cdot)$ interventions,
counterfactual queries, d-separation, and automatic identification of backdoor
and frontdoor adjustment sets.  The default bootstrap is Simpson's paradox; the
SCM grows during REM sleep via the PC algorithm applied to journal
observations.

\paragraph{VSA / HRR algebra.}
Holographic Reduced Representations encode triples as role/filler bundles via
circular convolution ($O(d \log d)$).  Unbinding recovers fillers via circular
correlation.  The codebook is persistent: the same atom name always produces the
same hypervector across runs.

\paragraph{Hopfield associative memory.}
Modern Continuous Hopfield networks \cite{Ramsauer2020} provide one-step
content-addressable retrieval with exponential storage capacity in the
embedding dimension~$d$.  The inverse temperature $\beta = \sqrt{d}/\sigma$ is
derived from the geometry of the stored patterns.

\paragraph{Conformal prediction.}
Split-conformal predictors (LAC and APS variants) wrap the LLM's softmax
outputs with a finite-sample coverage guarantee.  The size of the prediction set
$|C|$ serves as a Fristonian ambiguity signal: $|C| = 1$ collapses to a
confident answer; $|C| > 1$ raises an intrinsic cue that steers the LLM toward
a clarifying question.

\paragraph{Top-down cognitive control.}
Four predictive-coding-inspired mechanisms let the substrate override the LLM:
\begin{itemize}[nosep]
  \item \emph{Hypothesis masking:} Negative logit bias physically blocks
    rejected tokens; an iterative search prunes the hypothesis space until the
    LLM produces a valid answer.
  \item \emph{Epistemic interruption:} A streaming evaluator checks generated
    tokens every $k$ steps and, on a logical violation, truncates and
    re-injects a correction feature vector.
  \item \emph{Modality shifting:} A continuous unit-norm direction biases the
    residual stream toward a named cognitive mood (analytical, fluent, etc.).
  \item \emph{Causal constraint injection:} KV-memory grafts pre-loaded with
    $P(Y \mid \doop(T{=}t))$ pull the LLM toward the SCM's verdict whenever
    it attends to the cause concept.
\end{itemize}

\subsection{Evaluation Framework}

We evaluate on three tiers:

\begin{enumerate}[nosep]
  \item \textbf{Standard NLP benchmarks:} Multiple-choice tasks scored by
    length-normalized continuation log-likelihood (BoolQ, PIQA, ARC,
    WinoGrande, HellaSwag, CommonsenseQA, OpenBookQA, MMLU) and
    generation tasks scored by deterministic greedy decode with normalized
    exact matching (GSM8K).  A paired vanilla-vs-\emph{Broca shell} comparison
    verifies that the graft infrastructure does not degrade leaderboard
    accuracy.
  \item \textbf{Architecture probes:} Scripted prompts where the bare LLM and
    the full \emph{Broca shell} are scored on speech-exact and answer-present metrics.
    The probes cover semantic-memory recall, active-inference action selection,
    and causal-intervention queries.
  \item \textbf{Substrate benchmarks:} Eight CPU-only benchmarks that exercise
    each algebraic component directly: belief revision, hypothesis masking,
    Simpson's paradox resolution, memory fidelity, conformal coverage, VSA
    round-trip, Hopfield retrieval, and active-inference decision quality.
\end{enumerate}

All experiments are reproducible from a single command (\texttt{make paper})
and the results, tables, figures, and prose in Section~\ref{sec:experiments}
are regenerated automatically from the latest benchmark run.
