\section{Discussion}
\label{sec:discussion}

\subsection{Interpretation of Results}

The three evaluation tiers serve complementary purposes.  Standard NLP
benchmarks (Section~\ref{subsec:exp-hf-native}) establish that the graft
infrastructure does not degrade the frozen model's leaderboard accuracy---a
necessary sanity check before claiming any benefit from the substrate.  The
architecture probes (Section~\ref{subsec:exp-broca-arch}) demonstrate that the
grafts can steer the LLM toward the substrate's computed answers with high
fidelity, even when the bare model has no access to the relevant information.
The substrate benchmarks (Section~\ref{subsec:exp-substrate}) verify the
mathematical guarantees of each component in isolation.

\subsection{Why Grafts Instead of Prompts?}

The graft-based communication protocol has several advantages over
prompt-stuffing:

\begin{itemize}[nosep]
  \item \textbf{Constant token cost.}  Grafts do not consume context-window
    tokens.  The substrate can maintain arbitrarily large state (memory, world
    model, policies) without truncating the user's conversation.
  \item \textbf{Per-step influence.}  A prompt is read once at encoding time.
    A graft is applied at every decoding step, giving the substrate continuous
    influence over the generation trajectory.
  \item \textbf{Direct residual access.}  Grafts operate in the model's hidden
    geometry rather than in token space, enabling finer-grained control (e.g.\
    modality shifts, SNR-calibrated biasing).
  \item \textbf{Confidence modulation.}  Graft magnitude is annealed by
    substrate confidence, so the LLM retains autonomy when the substrate is
    uncertain.
\end{itemize}

\subsection{Limitations}

\paragraph{Scale.}
All experiments use Llama~3.2 1B Instruct.  Whether the graft protocol scales
to larger checkpoints (7B, 70B) without retuning the target SNR is an open
question.  The algebraic substrate components (VSA, Hopfield, SCM) are
model-size-agnostic, but the graft's residual-stream injection may need
recalibration for higher-dimensional hidden states.

\paragraph{Graft fidelity on open-ended tasks.}
Architecture probes use scripted, short-answer prompts where the expected output
is unambiguous.  On open-ended generation tasks (summarisation, creative
writing), the graft's influence is a soft bias, not a hard constraint.
Measuring graft effectiveness on such tasks requires evaluation metrics beyond
exact match.

\paragraph{SCM scalability.}
The finite SCM uses exact enumeration over exogenous worlds, which is
tractable only for toy cardinalities.  Real-world causal models will require
the Monte Carlo paths (\texttt{probability\_monte\_carlo},
\texttt{counterfactual\_probability\_monte\_carlo}) or approximate inference.

\paragraph{Conformal calibration in non-stationary settings.}
The conformal coverage guarantee holds under exchangeability of calibration
and test data.  In a conversational agent where the distribution shifts as the
user explores new topics, online recalibration (or a sliding-window conformal
predictor) is needed.

\subsection{Relationship to Prior Work}

The separation of language generation from cognitive control echoes classical
symbolic AI's division between a knowledge base and a natural-language
interface, but with a critical difference: here the ``interface'' is a full
transformer decoder, and the communication channel is continuous (residual
biasing) rather than symbolic (template filling).

The graft mechanism is related to activation engineering, representation
engineering, and steering vectors, but differs in that the bias is
\emph{per-step}, \emph{confidence-modulated}, and \emph{derived from a
structured world model} rather than from a contrastive pair of prompts.

Active inference over POMDPs for LLM control has been explored in the agent
literature, but typically as an outer loop that selects between prompt variants.
Our approach embeds the POMDP decision directly into the decoding loop via
graft biasing, avoiding the latency and token cost of re-prompting.
