% !TeX root = ../main.tex

\section{Conclusion}
\label{sec:conclusion}

We have presented \textsc{Mosaic}, an architecture that decouples language
generation from cognitive control by treating a frozen LLM as a language organ
biased through grafts by a persistent, mathematically grounded substrate.  The
three-tier evaluation framework---standard NLP benchmarks, architecture probes,
and substrate-specific benchmarks---provides complementary evidence that:

\begin{enumerate}[nosep]
  \item The graft infrastructure preserves the frozen model's leaderboard
    accuracy when no substrate signal is injected.
  \item The full substrate can steer the frozen decoder toward correct
    verbalisation of computed answers with high fidelity.
  \item Each algebraic component (belief revision, causal reasoning, VSA
    algebra, Hopfield retrieval, conformal prediction, active inference)
    satisfies its published mathematical guarantees.
\end{enumerate}

\paragraph{Future work.}
Several directions are open:
\begin{itemize}[nosep]
  \item Scaling the graft protocol to larger checkpoints (7B, 70B) and
    measuring whether SNR calibration transfers.
  \item Replacing the finite SCM with a learned causal model discovered from
    real conversation data (the PC-algorithm path already exists in the DMN's
    REM-sleep phase).
  \item Benchmarking on open-ended generation tasks with human evaluation of
    substrate-steered outputs.
  \item Integrating the self-improvement loop (Docker-based patch $\to$ test
    $\to$ PR $\to$ paper refresh) into a continuous evaluation pipeline.
  \item Extending the conformal predictor to a sliding-window variant for
    non-stationary conversational settings.
\end{itemize}

All code, benchmarks, and this paper are available at
\texttt{https://huggingface.co/theapemachine/mosaic}.  The benchmark-to-paper
pipeline ensures that every table, figure, and result paragraph in
Section~\ref{sec:experiments} reflects the most recent benchmark run.
