\subsection{Broca Architecture Probes}
\label{subsec:exp-broca-arch}

\paragraph{Protocol.}
We compare the bare frozen language host (\texttt{meta-llama/Llama-3.2-1B-Instruct}) against the full Broca cognitive substrate on scripted evaluation cases. Each case presents a prompt to both conditions and scores the output on two metrics: \emph{speech-exact} (verbatim match against the substrate's reference utterance) and \emph{answer-present} (the correct content word appears anywhere in the output). The baseline arm disables all grafts; the enhanced arm activates the full stack (semantic memory, active-inference agents, causal SCM, and all three graft types).

\begin{table}[htbp]
\centering
\caption{Architecture probe metrics: bare frozen LM vs.\ full Broca cognitive substrate.}
\label{tab:broca-arch-probes}
\begin{tabular}{lcc}
\toprule
Condition & Speech-exact & Answer-present \\
\midrule
Baseline (frozen LM) & $0.000$ & $0.000$ \\
Broca-enhanced & $0.000$ & $0.000$ \\
$\Delta$ (Broca $-$ baseline) & $0.000$ & $0.000$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Results.}
Table~\ref{tab:broca-arch-probes} compares the bare frozen language host (\texttt{meta-llama/Llama-3.2-1B-Instruct}) against the full Broca architecture on 2 scripted evaluation cases spanning semantic memory recall, active-inference action selection, and causal intervention queries. 
Under this snapshot, \emph{both} conditions obtain 0.0\% speech-exact accuracy and 0.0\% answer-present accuracy ($\Delta = 0.000$ speech-exact; $\Delta = 0.000$ answer-present), i.e., neither arm satisfied the scripted scoring criteria on these probes. This invites debugging (prompt formatting vs.\ reference strings, tokenizer alignment, or harness drift) rather than treating the tied zeros as comparable competence. 

