\subsection{Substrate-Specific Benchmarks}
\label{subsec:exp-substrate}

\paragraph{Protocol.}
We evaluate 8 capabilities that are unique to the cognitive substrate and not captured by standard NLP leaderboards. Each benchmark exercises a specific algebraic or control-theoretic guarantee: belief revision (rule-shift adaptation), top-down hypothesis masking (adversarial prompt resistance), do-calculus over a finite SCM (causal reasoning), SQLite-backed triple store fidelity, split-conformal prediction coverage, HRR bind/unbind algebra, Modern Continuous Hopfield retrieval, and EFE-driven POMDP decision quality. All benchmarks run on CPU without model downloads, exercising the substrate's algebra directly.

\begin{table}[htbp]
\centering
\caption{Substrate benchmark suite: per-benchmark scores and pass/fail status. \textit{Suite total}: the Pass column reports $n_{\mathrm{passed}}/n_{\mathrm{benchmarks}}$; the Score column is the arithmetic mean of the eight per-benchmark scores (not the pass rate).}
\label{tab:substrate-benchmarks}
\input{include/experiment/substrate_benchmark_table}
\end{table}

\paragraph{Results.}
The substrate benchmark suite exercises 8 capabilities that are orthogonal to standard NLP benchmarks. Of these, 8 pass their success criteria (pass rate: 100.0\%) in a total wall-clock time of 0.5s. We discuss each benchmark below.

\textit{Rule-shift adaptation.} We repeat the revision scenario 30 times (nominal 95\% CI $[1.0000, 1.0000]$, deterministic seeds: \texttt{trial\_seed = base\_seed + trial\_index * 1000003}); the aggregate mean score is 100.0\%. The belief-revision engine is seeded with 5 corroborating claims for ``\texttt{rome}'' and then presented with 8 low-surprise challenger claims for ``\texttt{paris}.'' After consolidation (representative episode), we observe 1 reflection(s), final log-odds (nats) $= 1.0012$, and a normalized completeness score $= 1.0000$ (relative to threshold 0.3). The stored belief successfully revises to the challenger value, confirming that the log-odds-based consolidation engine with prediction-gap-weighted trust correctly implements online Bayesian belief revision.

\textit{Adversarial prompt resistance.} We simulate 50 trials where 15 out of 100 tokens are designated as ``bad'' and the hypothesis masking graft must iteratively ban rejected candidates until a valid token is found. The search converges on a valid token in 100.0\% of trials with an average of 1.12 steps to convergence. 
This confirms that the iterative ban mechanism provides reliable top-down control.

\textit{Causal reasoning (Simpson's paradox).} The finite SCM correctly reproduces Simpson's paradox: naive association suggests treatment hurts ($P(Y{=}1 | T{=}1) = 0.3500$ vs.\ $P(Y{=}1 | T{=}0) = 0.6500$), while do-calculus reveals the true average treatment effect ATE $= 0.1000$ ($P(Y{=}1 | \mathrm{do}(T{=}1)) = 0.5500$, $P(Y{=}1 | \mathrm{do}(T{=}0)) = 0.4500$). 
The SCM's exact enumeration correctly recovers the interventional distribution.

\textit{Semantic memory fidelity.} We write 100 random (subject, predicate, object) triples to the SQLite-backed semantic memory and recall each. The recall rate is 100.0\% with mean confidence error $0$, confirming that the WAL-based storage engine preserves triple fidelity across the write-read cycle.

\textit{Conformal coverage guarantee.} We calibrate both LAC and APS conformal predictors on 200 synthetic distributions and evaluate on 500 held-out items at $\alpha = 0.1$ (target coverage $\geq 90.0\%$). Empirical coverage is 90.4\% (LAC) and 98.4\% (APS); the scalar headline score 94.4\% is their unweighted mean (formula in \texttt{score\_methodology} within the benchmark JSON). 
Both predictors meet the calibrated finite-sample coverage targets under our slack tolerance. Average prediction set sizes are 2.6 (LAC) and 3.52 (APS).

\textit{VSA algebraic fidelity.} We encode 150 random triples as HRR bundles via circular convolution and test role-unbinding accuracy across dimensionalities $d \in \{1000, 5000, 10000\}$. 
Unbinding accuracy: $d = 1000$: 100.0\%; $d = 5000$: 100.0\%; $d = 10000$: 100.0\%. 
Accuracy is at ceiling under this easy binding/unbinding regime, so dimensional scaling does not yet separate---the theoretical capacity curve $\sim 0.5 \cdot d / \log d$ would appear only under harder bundles or noise. 

\textit{Hopfield retrieval.} We store varying numbers of random unit-norm patterns in a Modern Continuous Hopfield network ($d = 256$) and query with noisy probes ($\sigma = 0.3$). 
Retrieval accuracy (cosine $> 0.8$): $N = 10$: 100.0\%; $N = 50$: 72.0\%; $N = 100$: 84.0\%; $N = 500$: 52.0\%. 
Accuracy vs.\ store size need not be monotone: probes inject Gaussian noise ($\sigma = 0.3$) and each $N$ uses fresh draws, so finite-sample interference dominates short rankings \cite{Ramsauer2020}. 

\textit{Active-inference decision quality.} The EFE-driven Tiger POMDP agent is evaluated over 200 episodes (max 3 steps each) against a uniform random baseline. The agent achieves a success rate of 87.0\% vs.\ 42.5\% for random (advantage: +0.4450; mean return: 0.6000 vs.\ -0.6508). 
The positive advantage confirms that Expected Free Energy minimization produces systematically better decisions than random exploration.


